---
title: "Combined AFM and Weekly Data"
author: "Data Analysis Report"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Data Loading and Setup

```{r}
DATA_PATH    <- "All_Data_1884_2013_0215_193821.csv"
AFM_DIR      <- "afm_outputs_what"
OUTPUT_FILE  <- "student_week_aggregations_what.csv"
```

# Key variables

```{r}
# Control how duration is calculated
USE_PROVIDED_DURATION <- TRUE  # TRUE: use step_duration_sec, FALSE: calculate from step_end_time - step_start_time

# Control how proficiency is calculated
PROFICIENCY_METHOD <- "new_skills_mastered"  # "mean": average proficiency across all skills
                              # "new_skills_mastered": count of NEW skills reaching >95% proficiency each week
# the two methods are intended for different use cases

# Control model fitting
USE_CACHED_MODEL <- FALSE  # Set to TRUE to load existing model instead of refitting

# Filter to last N weeks (set to NULL to use all data)
LAST_N_WEEKS <- 4  # Examples: NULL (all data), 52 (last year), 12 (last quarter), 4 (last month)
```

# Libraries

```{r}
suppressPackageStartupMessages({
  library(readr)      # fast CSV I/O
  library(dplyr)      # data wrangling
  library(tidyr)      # separate_rows()
  library(janitor)    # clean_names()
  library(lme4)       # glmer()
  library(fs)         # file system helpers
  library(stringr)    # string manipulation
  library(tibble)     # tibble utilities
  library(lubridate)  # date functions like ymd_hms, isoyear, isoweek
})

source("data_cleaning_helpers.R")
```

# 1. Load data

```{r}
raw <- read_delim(DATA_PATH, delim = "\t", show_col_types = FALSE) |>
         clean_names()

# Print sample of raw data
print("Sample of raw data:")
raw |> head(5) |> knitr::kable()
```

# 2. Parse dates and create week identifiers

```{r}
d_with_week_id <- raw %>%
  mutate(
    step_start_datetime = ymd_hms(step_start_time),
    duration_sec = if (USE_PROVIDED_DURATION) {
      as.numeric(step_duration_sec)
    } else {
      step_end_datetime <- ymd_hms(step_end_time)
      as.numeric(difftime(step_end_datetime, step_start_datetime, units = "secs"))
    },

    # Create week identifier using ISO year and week: YYYY-WNN, 2011-W37
    week_year = isoyear(step_start_datetime),
    week_num = isoweek(step_start_datetime),
    week_id = paste0(week_year, "-W", sprintf("%02d", week_num)),

  ) %>%
  # Apply filtering for valid timestamps
  filter(!is.na(step_start_datetime)) %>%
  log_filter_result(raw, ., "After filtering invalid timestamps")

# Apply time window filtering if requested
if (!is.null(LAST_N_WEEKS)) {
  # Find the cutoff week
  all_weeks <- unique(d_with_week_id$week_id) %>% sort()
  n_total_weeks <- length(all_weeks)
  
  if (LAST_N_WEEKS >= n_total_weeks) {
    cat("Note: Requested", LAST_N_WEEKS, "weeks, but data only contains", n_total_weeks, "weeks. Using all data.\n")
  } else {
    cutoff_week <- all_weeks[n_total_weeks - LAST_N_WEEKS + 1]
    n_before <- nrow(d_with_week_id)
    
    d_with_week_id <- d_with_week_id %>%
      filter(week_id >= cutoff_week)
    
    n_after <- nrow(d_with_week_id)
    n_students_after <- n_distinct(d_with_week_id$anon_student_id)
    
    cat("\n=== Time Window Filtering Applied ===\n")
    cat("Using last", LAST_N_WEEKS, "weeks of data\n")
    cat("Date range:", cutoff_week, "to", max(all_weeks), "\n")
    cat("Rows:", n_before, "->", n_after, "(", round(100 * n_after / n_before, 1), "% retained)\n")
    cat("Students in filtered data:", n_students_after, "\n")
    cat("=====================================\n\n")
  }
}

# Print sample of weekly data
cat("Sample of data after all filtering:\n")
d_with_week_id |> head(5) |> knitr::kable() |> print()
```

# 3. Process Skills/KCs

```{r}
# Process skills from the original weekly data, not the aggregated data
d_with_week_id_and_sep_skills <- d_with_week_id %>%
  separate_longer_delim(c(kc_sub_skills, opportunity_sub_skills), delim = "~~") %>%
  clean_skills_systematic("kc_sub_skills", min_length = 10, verbose = FALSE)

# print sample of student-week clean skills
d_with_week_id_and_sep_skills |> head(5) |> knitr::kable()
```

# 4. build AFM variables

```{r}
d_afm <- d_with_week_id_and_sep_skills %>%
  transmute(
    anon_student_id = anon_student_id,
    week_id = week_id,  # KEEP week_id for later use
    kc_sub_skills = kc_sub_skills,
    n_opportunity = as.integer(opportunity_sub_skills),
    outcome_bin = case_when(
      first_attempt == "correct" ~ 1L,
      first_attempt == "incorrect" ~ 0L,
      first_attempt == "hint" ~ 0L
    )
  ) %>%
  filter(!is.na(kc_sub_skills) & kc_sub_skills != "" & !is.na(outcome_bin))

# print sample of data changed to afm-appropriate formats
d_afm |> head(5) |> knitr::kable()
```

# 5. Fit AFM model with individual learning rates

```{r}
model_path <- path(AFM_DIR, "afm_model.rds")

if (USE_CACHED_MODEL && file_exists(model_path)) {
  cat("Loading cached AFM model from:", model_path, "\n")
  m_afm <- readRDS(model_path)
  cat("Model loaded successfully\n")
} else {
  cat("Fitting AFM model (this may take several minutes)...\n")
  if (!is.null(LAST_N_WEEKS)) {
    cat("📅 Using data from last", LAST_N_WEEKS, "weeks only\n")
  }
  cat("Data size:", nrow(d_afm), "observations\n")
  cat("Number of students:", n_distinct(d_afm$anon_student_id), "\n")
  cat("Number of skills:", n_distinct(d_afm$kc_sub_skills), "\n")
  cat("Week range:", min(d_afm$week_id), "to", max(d_afm$week_id), "\n\n")
  
  m_afm <- glmer(
    outcome_bin ~ n_opportunity + (1 + n_opportunity | anon_student_id) + (1 | kc_sub_skills),
    data = d_afm,
    family = binomial,
    nAGQ = 0,
    verbose = 2,  # Show iteration progress (0=none, 1=minimal, 2=detailed)
    control = glmerControl(
      optimizer = "bobyqa",
      calc.derivs = FALSE,
      check.conv.grad = .makeCC("warning", tol = 1e-3, relTol = NULL),
      check.conv.singular = .makeCC(action = "ignore", tol = 1e-3),
      check.conv.hess = .makeCC(action = "ignore", tol = 1e-3),
      optCtrl = list(maxfun = 100000)  # Max iterations before giving up
    )
  )
  
  cat("Model fitting complete\n")
  
  # Check for convergence warnings
  if (length(m_afm@optinfo$conv$lme4$messages) > 0) {
    cat("⚠️ Model convergence warnings:\n")
    print(m_afm@optinfo$conv$lme4$messages)
  }
}

# Print model summary
cat("\nModel summary:\n")
print(summary(m_afm))
``` 

# 6. extract parameters as latent features

```{r}

# TODO: check if implementation is correct

# Student abilities and individual learning rates
student_effects <- ranef(m_afm)$anon_student_id
student_abilities <- tibble(
  anon_student_id = rownames(student_effects),
  ability = student_effects[,1],
  learning_rate = student_effects[,2]
)

# Skill easiness
skill_effects <- ranef(m_afm)$kc_sub_skills
skill_easyness <- tibble(
  kc_sub_skills = rownames(skill_effects),
  easyness = skill_effects[,1]
)

# Global parameters
fixed_effects <- fixef(m_afm)
global_learning_rate <- fixed_effects["n_opportunity"]
intercept <- fixed_effects["(Intercept)"]
```

# 7. save outputs

```{r}

# TODO: check if this is correct
# Fixed: Using AFM_DIR instead of undefined OUTPUT_DIR
# Also fixed: global_learning_rate and intercept are scalars, need to be wrapped in tibbles

# Save both extracted parameters (as CSVs for easy inspection) and full model (as RDS for later use)
write_csv(student_abilities, path(AFM_DIR, "student_ability.csv"))
write_csv(skill_easyness, path(AFM_DIR, "skill_easyness.csv"))
write_csv(tibble(parameter = "n_opportunity", estimate = global_learning_rate), 
          path(AFM_DIR, "global_learning_rate.csv"))
write_csv(tibble(parameter = "intercept", estimate = intercept), 
          path(AFM_DIR, "intercept.csv"))

# Save the full model object for later use
saveRDS(m_afm, file = path(AFM_DIR, "afm_model.rds"))
cat("Saved AFM model to:", path(AFM_DIR, "afm_model.rds"), "\n")
```

# 8. calculate "proficiency" for each student-skill-week
This is done through looking at the last opportunity for each student-skill in each week

```{r}

# Calculate last opportunity for each student-skill-week combination
# Using d_afm directly since it now includes week_id

last_opportunities <- d_afm %>%
  group_by(anon_student_id, week_id, kc_sub_skills) %>%
  summarise(
    last_opportunity = max(n_opportunity, na.rm = TRUE),
    .groups = "drop"
  )

# print sample of last opportunities
last_opportunities |> head(5) |> knitr::kable()

# Create prediction data frame with the last opportunity for each student-skill-week
prediction_data <- last_opportunities %>%
  rename(n_opportunity = last_opportunity)  # predict() expects the same variable names as the model

# Use native predict() function to get proficiency estimates
proficiency_estimates <- prediction_data %>%
  mutate(
    proficiency = predict(m_afm, newdata = ., type = "response", allow.new.levels = TRUE)
  ) %>%
  rename(
    last_opportunity = n_opportunity  # Rename back for clarity
  )

# print sample of proficiency estimates
proficiency_estimates |> head(5) |> knitr::kable()
```

# 9. Create behavioral aggregations and calculate proficiency metric

First, create the student-week behavioral metrics that we'll need for the final output:
The decision is to use the original data, not continuing from the cleaned skills data.
The original thought was because when we filter out rows with suspicious skills,
we might also lose the time spent or problems solved on those skills, which didn't sound right.
Now I'm wondering if this mismatch between the skills and time spent on themis problematic...

```{r}
# Create behavioral aggregations (moved from block 3)
student_week_basic <- d_with_week_id %>%
  group_by(anon_student_id, week_id) %>%
  reframe(
    # Number of minutes per week (convert from seconds, handle NAs)
    minutes_per_week = sum(duration_sec, na.rm = TRUE) / 60,

    # Problems solved (count unique problems)
    problems_solved = n_distinct(problem_name),

    # Total opportunities (attempts at problem steps)
    total_opportunities = n()
  )

# print sample of student-week aggregated data
cat("Student-week behavioral metrics:\n")
student_week_basic |> head(5) |> knitr::kable() |> print()
```

Now calculate the proficiency metric based on the chosen method:

```{r}
# Calculate proficiency metric

if (PROFICIENCY_METHOD == "mean") {
  # Method 1: Average proficiency across all skills
  avg_proficiency <- proficiency_estimates %>%
    group_by(anon_student_id, week_id) %>%
    summarise(
      avg_proficiency = mean(proficiency, na.rm = TRUE),
      n_skills_measured = n(),
      .groups = "drop"
    )
  cat("Calculated average proficiency across skills\n")
} else if (PROFICIENCY_METHOD == "new_skills_mastered") {
  # Method 2: Count of NEW skills reaching >95% proficiency each week
  
  # Step 1: For each student-skill, find the first week they reached >95% proficiency
  first_mastery_week <- proficiency_estimates %>%
    filter(proficiency > 0.95) %>%
    group_by(anon_student_id, kc_sub_skills) %>%
    summarise(
      first_mastery_week = min(week_id),
      .groups = "drop"
    )
  
  # Step 2: Count how many skills were newly mastered in each week
  new_skills_per_week <- first_mastery_week %>%
    group_by(anon_student_id, first_mastery_week) %>%
    summarise(
      new_skills_mastered = n(),
      .groups = "drop"
    ) %>%
    rename(week_id = first_mastery_week)
  
  # Step 3: Create full student-week grid and fill in zeros for weeks with no new masteries
  all_student_weeks <- student_week_basic %>%
    distinct(anon_student_id, week_id)
  
  avg_proficiency <- all_student_weeks %>%
    left_join(new_skills_per_week, by = c("anon_student_id", "week_id")) %>%
    mutate(
      avg_proficiency = replace_na(new_skills_mastered, 0),
      n_skills_measured = NA_integer_  # not applicable for this counting method
    ) %>%
    select(-new_skills_mastered)
  
  cat("Calculated count of newly mastered skills per week\n")
} else {
  stop("Invalid PROFICIENCY_METHOD. Use 'mean' or 'new_skills_mastered'")
}

# print sample of proficiency metric
avg_proficiency |> head(5) |> knitr::kable()
```

# 10. combine all metrics

```{r}

# TODO: check if this is correct
final_aggregation <- student_week_basic %>%
  select(anon_student_id, week_id, minutes_per_week, problems_solved, total_opportunities) %>%
  left_join(avg_proficiency, by = c("anon_student_id", "week_id")) %>%
  # pretty sure I designed week_id to follow chronologically when sorted lexicographically
  arrange(anon_student_id, week_id)

  # Summary statistics
cat("\nSummary of student-week aggregations:\n")
cat("  Total student-weeks:", nrow(final_aggregation), "\n")
cat("  Unique students:", n_distinct(final_aggregation$anon_student_id), "\n")
cat("  Date range:", min(final_aggregation$week_id), "to", max(final_aggregation$week_id), "\n")
cat("\nMetric summaries:\n")
print(summary(select(final_aggregation, minutes_per_week, problems_solved, 
                    total_opportunities, avg_proficiency)))
```

# 11. save output

```{r}
write_csv(final_aggregation, OUTPUT_FILE)

cat("Saved output to", OUTPUT_FILE, "\n")
```





