---
title: "Student Efficiency and Outlier Analysis"
author: "Analysis Report"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 8)
```

## Overview

This analysis investigates students who complete problems unusually quickly and identifies potential gaming behavior versus legitimate high performers ("efficiency beasts").

```{r load-libraries}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)

# Load data
data <- read_csv("student_week_aggregations2.csv", show_col_types = FALSE)
```

## Data Preparation

```{r calculate-efficiency}
# Calculate efficiency metrics
efficiency_data <- data %>%
  filter(problems_solved > 0, minutes_per_week > 0) %>%
  mutate(
    seconds_per_problem = (minutes_per_week * 60) / problems_solved,
    problems_per_minute = problems_solved / minutes_per_week,
    seconds_per_opportunity = (minutes_per_week * 60) / total_opportunities,
    opportunities_per_problem = total_opportunities / problems_solved
  )

cat("Total student-weeks analyzed:", nrow(efficiency_data), "\n")
```

## 1. Speed Demons Analysis

Identifying students who complete problems in less than 30 seconds.

```{r speed-demons}
speed_demons <- efficiency_data %>%
  filter(seconds_per_problem < 30) %>%
  arrange(seconds_per_problem)

cat("Number of student-weeks with <30 sec/problem:", nrow(speed_demons), "\n")
cat("Unique students:", n_distinct(speed_demons$anon_student_id), "\n")
cat("Percentage of all student-weeks:", 
    round(nrow(speed_demons) / nrow(efficiency_data) * 100, 2), "%\n")
```

### Top 10 Fastest Problem Solvers

```{r fastest-students}
speed_demons %>%
  select(anon_student_id, week_id, seconds_per_problem, problems_solved, 
         avg_proficiency, opportunities_per_problem) %>%
  slice_head(n = 10) %>%
  kable(digits = 2, 
        col.names = c("Student ID", "Week", "Seconds/Problem", 
                      "Problems Solved", "Avg Proficiency", "Opportunities/Problem"))
```

## 2. Proficiency vs Speed Analysis

Examining the relationship between solving speed and learning outcomes.

```{r speed-groups}
speed_groups <- efficiency_data %>%
  mutate(
    speed_category = case_when(
      seconds_per_problem < 30 ~ "Very Fast (<30s)",
      seconds_per_problem < 60 ~ "Fast (30-60s)",
      seconds_per_problem < 120 ~ "Normal (60-120s)",
      TRUE ~ "Slow (>120s)"
    )
  ) %>%
  group_by(speed_category) %>%
  summarise(
    n = n(),
    avg_proficiency = mean(avg_proficiency, na.rm = TRUE),
    median_proficiency = median(avg_proficiency, na.rm = TRUE),
    avg_problems = mean(problems_solved),
    avg_opportunities_per_problem = mean(opportunities_per_problem)
  ) %>%
  arrange(avg_proficiency)

kable(speed_groups, digits = 3,
      col.names = c("Speed Category", "Count", "Avg Proficiency", 
                    "Median Proficiency", "Avg Problems", "Avg Opportunities/Problem"))
```

### Proficiency vs Speed Visualization

```{r proficiency-speed-plot, fig.height=6}
ggplot(efficiency_data, aes(x = seconds_per_problem, y = avg_proficiency)) +
  geom_point(alpha = 0.3) +
  scale_x_log10() +
  geom_smooth(method = "loess", color = "red") +
  geom_vline(xintercept = 30, color = "red", linetype = "dashed") +
  labs(
    title = "Proficiency vs Problem Solving Speed",
    subtitle = "Red line at 30 seconds per problem threshold",
    x = "Seconds per Problem (log scale)",
    y = "Average Proficiency"
  ) +
  theme_minimal()
```

## 3. Weekly Patterns

Identifying weeks with unusual activity patterns.

### Weeks with Fastest Average Completion Times

```{r weekly-patterns}
weekly_stats <- efficiency_data %>%
  group_by(week_id) %>%
  summarise(
    n_students = n(),
    avg_seconds_per_problem = mean(seconds_per_problem),
    median_seconds_per_problem = median(seconds_per_problem),
    total_problems = sum(problems_solved),
    avg_proficiency = mean(avg_proficiency)
  ) %>%
  arrange(avg_seconds_per_problem)

weekly_stats %>% 
  slice_head(n = 10) %>%
  kable(digits = 1,
        col.names = c("Week", "Students", "Avg Sec/Problem", 
                      "Median Sec/Problem", "Total Problems", "Avg Proficiency"))
```

### Weeks with Most Activity

```{r high-activity-weeks}
weekly_stats %>% 
  arrange(desc(total_problems)) %>% 
  slice_head(n = 10) %>%
  kable(digits = 1,
        col.names = c("Week", "Students", "Avg Sec/Problem", 
                      "Median Sec/Problem", "Total Problems", "Avg Proficiency"))
```

## 4. Potential Gaming Behavior

Identifying students who are very fast but have low proficiency, suggesting potential gaming of the system.

```{r gaming-suspects}
gaming_suspects <- efficiency_data %>%
  filter(
    seconds_per_problem < 30,
    avg_proficiency < 0.6,
    problems_solved > 20  # Significant activity
  ) %>%
  arrange(seconds_per_problem)

cat("Students with <30s/problem AND <0.6 proficiency AND >20 problems:", nrow(gaming_suspects), "\n")
cat("Percentage of speed demons:", 
    round(nrow(gaming_suspects) / nrow(speed_demons) * 100, 2), "%\n")
```

### Top Gaming Suspects

```{r gaming-table}
gaming_suspects %>%
  select(anon_student_id, week_id, seconds_per_problem, problems_solved, 
         avg_proficiency, opportunities_per_problem) %>%
  slice_head(n = 10) %>%
  kable(digits = 2,
        col.names = c("Student ID", "Week", "Seconds/Problem", 
                      "Problems Solved", "Avg Proficiency", "Opportunities/Problem"))
```

## 5. Efficiency Beasts

Students who solve many problems quickly while maintaining high proficiency - the legitimate high performers.

```{r efficiency-beasts}
efficiency_beasts <- efficiency_data %>%
  filter(
    problems_solved > 100,
    avg_proficiency > 0.7,
    seconds_per_problem < 60
  ) %>%
  arrange(desc(problems_solved))

cat("Students solving >100 problems with >0.7 proficiency in <60s/problem:", nrow(efficiency_beasts), "\n")
```

### Top Efficiency Beasts

```{r beasts-table}
efficiency_beasts %>%
  select(anon_student_id, week_id, problems_solved, seconds_per_problem,
         avg_proficiency, minutes_per_week) %>%
  slice_head(n = 10) %>%
  kable(digits = 2,
        col.names = c("Student ID", "Week", "Problems Solved", 
                      "Seconds/Problem", "Avg Proficiency", "Minutes/Week"))
```

## 6. Efficiency Distribution

### Problem Solving Speed Distribution

```{r efficiency-distribution, fig.height=6}
ggplot(efficiency_data, aes(x = seconds_per_problem)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  scale_x_log10() +
  geom_vline(xintercept = c(30, 60, 120), color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Problem Solving Speed",
    subtitle = "Red lines at 30s, 60s, 120s thresholds",
    x = "Seconds per Problem (log scale)",
    y = "Count"
  ) +
  theme_minimal()
```

## 7. Statistical Summary

### Overall Efficiency Statistics

```{r statistics}
# Summary statistics
summary_stats <- summary(efficiency_data$seconds_per_problem)
print(summary_stats)

# Correlation analysis
cor_test <- cor.test(log(efficiency_data$seconds_per_problem), 
                     efficiency_data$avg_proficiency)

cat("\nCorrelation between log(speed) and proficiency:\n")
cat("  Correlation coefficient:", round(cor_test$estimate, 3), "\n")
cat("  P-value:", format(cor_test$p.value, scientific = TRUE), "\n")
cat("  95% CI: [", round(cor_test$conf.int[1], 3), ",", 
    round(cor_test$conf.int[2], 3), "]\n")
```

### Key Findings Summary

```{r findings-summary}
findings <- data.frame(
  Metric = c(
    "Total student-weeks analyzed",
    "Speed demons (<30s/problem)",
    "Potential gaming instances",
    "Efficiency beasts (high volume + high proficiency)",
    "Correlation (log speed vs proficiency)",
    "Week with most gaming suspects(2012-W21)"
  ),
  Value = c(
    nrow(efficiency_data),
    nrow(speed_demons),
    nrow(gaming_suspects),
    nrow(efficiency_beasts),
    round(cor_test$estimate, 3),
    sum(gaming_suspects$week_id == "2012-W21")
  )
)

kable(findings, col.names = c("Metric", "Value"))
```

## Conclusions

1. **Speed-Proficiency Relationship**: There is a positive correlation (r = `r round(cor_test$estimate, 3)`) between log(solving time) and proficiency, indicating that students who take more time per problem tend to achieve better learning outcomes.

2. **Gaming Behavior**: Approximately `r round(nrow(gaming_suspects) / nrow(efficiency_data) * 100, 2)`% of student-weeks show potential gaming behavior (very fast solving with low proficiency).

3. **Efficiency Beasts**: `r nrow(efficiency_beasts)` student-weeks demonstrate exceptional performance - solving over 100 problems with high proficiency at reasonable speeds.

4. **Week 2012-W21 Anomaly**: This week shows a concentrated pattern of gaming behavior with many students solving exactly 44 problems at impossibly fast speeds.

5. **Data Quality**: The presence of students solving problems in 1-2 seconds suggests either system gaming, or data collection issues that should be investigated further. 