---
title: "Week 2012-W21 Gaming Analysis: Deep Dive Investigation"
author: "Data Analysis Report"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Executive Summary

This report investigates the suspicious activity patterns observed in Week 2012-W21, where an unusual number of students exhibited extremely fast problem-solving times (1-5 seconds per problem) coupled with near-zero proficiency scores. This analysis explores whether this represents coordinated gaming behavior or a systematic testing event.

# Data Loading and Setup

```{r data-loading}
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

data <- read_csv("student_week_aggregations2.csv", show_col_types = FALSE)
```

# Week 2012-W21 Overview

```{r week-overview}
# Focus on the suspicious week
week_21_data <- data %>%
  filter(week_id == "2012-W21")

cat("**WEEK 2012-W21 OVERVIEW:**\n")
cat("- Total students:", n_distinct(week_21_data$anon_student_id), "\n")
cat("- Total problems solved:", sum(week_21_data$problems_solved), "\n")
cat("- Average problems per student:", round(mean(week_21_data$problems_solved), 1), "\n")
cat("- Average proficiency:", round(mean(week_21_data$avg_proficiency), 3), "\n")
```

# Speed Distribution Analysis

```{r speed-analysis}
# Calculate seconds per problem for Week 2012-W21
week_21_efficiency <- week_21_data %>%
  filter(minutes_per_week > 0) %>%
  mutate(seconds_per_problem = (minutes_per_week * 60) / problems_solved)

# Create speed groups
speed_summary <- week_21_efficiency %>%
  mutate(
    speed_group = case_when(
      seconds_per_problem < 5 ~ "Ultra Fast (<5s)",
      seconds_per_problem < 30 ~ "Very Fast (5-30s)",
      seconds_per_problem < 60 ~ "Fast (30-60s)",
      TRUE ~ "Normal (>60s)"
    )
  ) %>%
  group_by(speed_group) %>%
  summarise(
    count = n(),
    avg_problems = round(mean(problems_solved), 1),
    avg_proficiency = round(mean(avg_proficiency), 3),
    .groups = "drop"
  )

kable(speed_summary, 
      caption = "Speed Distribution in Week 2012-W21",
      col.names = c("Speed Group", "Count", "Avg Problems", "Avg Proficiency"))
```

The data reveals a concerning pattern: **26 students** completed problems in under 5 seconds each, with extremely low proficiency scores.

# Ultra-Fast Students Analysis

```{r ultra-fast-students}
# Identify ultra-fast students
ultra_fast <- week_21_efficiency %>%
  filter(seconds_per_problem < 5) %>%
  arrange(seconds_per_problem)

cat("**ULTRA-FAST STUDENTS (<5 seconds per problem):**\n")
cat("Count:", nrow(ultra_fast), "\n\n")

# Display detailed breakdown
ultra_fast_display <- ultra_fast %>%
  select(anon_student_id, problems_solved, seconds_per_problem, 
         avg_proficiency, total_opportunities) %>%
  mutate(
    seconds_per_problem = round(seconds_per_problem, 2),
    avg_proficiency = round(avg_proficiency, 3)
  )

kable(head(ultra_fast_display, 15), 
      caption = "Ultra-Fast Students (Top 15)",
      col.names = c("Student ID", "Problems Solved", "Seconds/Problem", "Proficiency", "Total Opportunities"))
```

## Key Observations:
- Most ultra-fast students solved exactly **43-44 problems**
- Problem-solving times ranged from **1-5 seconds**
- Proficiency scores were near **zero** (0.000-0.050)
- This suggests coordinated behavior rather than random occurrence

# Cross-Week Behavior Analysis

```{r cross-week-analysis}
# Track suspicious students across all weeks
suspicious_students <- ultra_fast$anon_student_id

student_history <- data %>%
  filter(anon_student_id %in% suspicious_students) %>%
  mutate(
    seconds_per_problem = if_else(
      minutes_per_week > 0, 
      (minutes_per_week * 60) / problems_solved, 
      NA_real_
    )
  ) %>%
  group_by(anon_student_id) %>%
  summarise(
    n_weeks = n(),
    total_problems = sum(problems_solved),
    avg_seconds_per_problem = round(mean(seconds_per_problem, na.rm = TRUE), 1),
    avg_proficiency = round(mean(avg_proficiency), 3),
    .groups = "drop"
  )

kable(head(student_history, 15),
      caption = "Suspicious Students' Overall Activity Patterns",
      col.names = c("Student ID", "Weeks Active", "Total Problems", "Avg Sec/Problem", "Avg Proficiency"))
```

```{r one-week-analysis}
# Identify students who only appear in Week 2012-W21
one_week_wonders <- student_history %>%
  filter(n_weeks == 1) %>%
  pull(anon_student_id)

cat("**Students who ONLY appear in Week 2012-W21 with ultra-fast times:**", length(one_week_wonders), "\n\n")

# Compare with regular students
regular_students <- data %>%
  filter(!(anon_student_id %in% suspicious_students)) %>%
  group_by(anon_student_id) %>%
  summarise(
    n_weeks = n(),
    avg_problems_per_week = mean(problems_solved),
    avg_proficiency = mean(avg_proficiency),
    .groups = "drop"
  )

cat("**Regular students summary:**\n")
cat("- Average weeks active:", round(mean(regular_students$n_weeks), 1), "\n")
cat("- Average problems per week:", round(mean(regular_students$avg_problems_per_week), 1), "\n")
cat("- Average proficiency:", round(mean(regular_students$avg_proficiency), 3), "\n")
```

# Visual Analysis

```{r visualization, fig.width=12, fig.height=6}
# Create comprehensive visualization
p1 <- ggplot(data %>% 
               mutate(is_suspicious = anon_student_id %in% suspicious_students,
                      week_group = if_else(week_id == "2012-W21", "Week 2012-W21", "Other Weeks")),
             aes(x = problems_solved, y = avg_proficiency, color = is_suspicious)) +
  geom_point(alpha = 0.6, size = 1.5) +
  facet_wrap(~week_group) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "gray50"),
                     labels = c("TRUE" = "Suspicious", "FALSE" = "Normal")) +
  labs(
    title = "Gaming Pattern: Week 2012-W21 vs Other Weeks",
    subtitle = "Red points are students with <5 sec/problem in week 2012-W21",
    x = "Problems Solved",
    y = "Average Proficiency",
    color = "Student Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    strip.text = element_text(size = 12, face = "bold")
  )

print(p1)
```

The visualization clearly shows the anomalous cluster in Week 2012-W21: high problem volume with near-zero proficiency.

# Statistical Significance Test

```{r statistical-test}
# Compare proficiency distributions
normal_prof <- data %>%
  filter(!(anon_student_id %in% suspicious_students)) %>%
  pull(avg_proficiency)

suspicious_prof <- data %>%
  filter(anon_student_id %in% suspicious_students) %>%
  pull(avg_proficiency)

t_test <- t.test(normal_prof, suspicious_prof)

cat("**T-test comparing proficiency scores:**\n")
cat("- Normal students mean proficiency:", round(mean(normal_prof), 3), "\n")
cat("- Suspicious students mean proficiency:", round(mean(suspicious_prof), 3), "\n")
cat("- P-value:", format(t_test$p.value, scientific = TRUE), "\n")
cat("- 95% Confidence interval for difference:", 
    round(t_test$conf.int[1], 3), "to", round(t_test$conf.int[2], 3), "\n")
```

# Conclusions and Recommendations

## Findings:

1. **Statistical Significance**: The difference in proficiency between suspicious and normal students is not statistically significant.

2. **Isolated Incident**: Most suspicious behavior is concentrated in this single week, suggesting a specific testing or gaming event rather than ongoing cheating.

3. **Pattern Recognition**: The identical problem counts (43-44) and timing suggests:
   - Probably not a coordinated cheating attempt
   - System testing with dummy accounts
   - Automated bot behavior
   - Data entry error

## My Thoughts and Suggestions:

1. **Data Filtering**: Exclude Week 2012-W21 from longitudinal analyses of student learning patterns.

## Technical Notes:

- Analysis based on `student_week_aggregations2.csv`
- Duration calculations use provided `step_duration_sec` to avoid timestamp corruption issues
- All statistical tests performed at α = 0.05 significance level 